Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-23 manuscript/sa_construction.tex
finished pfg
finished iterator preparation

\section{Suffix array construction}
Suffix array is a permutation of string positions which lexicographically
sorts the suffixes of the string starting at that position.
It is extremely powerful data structure with many applications in efficient
algorithms on strings.
Enhanced by additional data structures, such as for example longest common
prefix (LCP) array, it can fully substitute suffix tree and solve problems,
such as exact pattern matching, repeat finding, maximum exact matching (MEM)
finding, document retrieval and many more.

There exists several algorithms for suffix array construction in linear time
\cite{} with several practical implementations \cite{}.
Altough their time complexity is linear, these algorithms become bottlenecks in
some application due to their linear space complexity.
This is especially relevant in pangenomics, where the datasets often do not fit
in the memory.
Nonetheless, as we shown in previous sections, prefix-free graphs can represent
the pangenome in substantialy smaller space and therefore can fit in computer
memory.

Here we show another crucial advantage of prefix-free graphs - their ability to
output the suffix array of the original string iterativelly.
This allows us to generate the suffix array one-by-one with the possibility
to compress or use in subsequential computations directly, giving us the ability
to use the rich theory of suffix arrays without ever storing them.

\subsection{Iterator preparation}
In order to prepare the prefix-free graph to generate the suffix array of
original pangenome we need to create several data structures.
First, we concatenate all the prefix-free segments into a single string using
a separator $\#$ and append a sentinel $\$$.
We will call this concatenation \emph{segment join}.
An example of a segment join is in the Figure \ref{fig:ids_and_positions}.

Next, for the segment join we calculate the suffix array and the LCP array.
For this we can use the linear time, linear space algorithms mentioned previously,
since the segment join is usually much smaller than the original pangenome.

Next, for each suffix of a segment join, we need to calculate the corresponding
segment ID and the segment position.
These two arrays represent the current suffix of a segment join starts with a
particular segment and the position in this segment.
This can be computed using an inverse permutation of a suffix array of a segment
join in linear time.
To illustrate the procedure, consider the segment join of our running example from
Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in current
segment by linearly iterating the segment join and incrementing ID and position
accordingly.
Then, applying the inverse permutation of a suffix array to these arrays will
change the order of computed values in a way corresponding to the sorted suffixes.
The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
segment ID array and segment position array.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
    \caption{
        Segment IDs and segment positions of a segment join. Permuting these
        according to the inverse permutation of a suffix array results in 
        segment ID array and segment position array shown in the Table \ref{fig:segment_sa}.
    }
    \label{fig:ids_and_positions}
\end{figure}

@@

In the Table \ref{tab:suffix}, one row can represent multiple positions of the
pangenome.
To identify these positions, we store some additional information in a second
table \ref{tab:segment}.
For each segment, we store it's length, it's starting positions in the pangenome
and the rank of the right context of these positions.
To calculate the starting positions and the ranks of the right contexts we use
a path join.
Similarly to segment join, a path join is a concatenation with delimiters $\#$
and sentinel $\$$, but now constructed by concatenating the paths.
An example of a path join for our running example is in the Figure \ref{fig:path_join}.
Starting positions can then be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.
Computation of the ranks is more involved.
We need to construct the suffix array of the path join and find it's inverse
permutation $ISA$.
This gives us ranks for each position in the path join.
To determine the rank of the right context for position $i$, we simply take
the value of the ranks $ISA[i+1]$.
Finaly, we store the starts and ranks sorted by the rank values in the segment
table as shown in Table \ref{tab:segment}.

@@

>>>>

\section{Suffix array construction}
A suffix array is a permutation of string positions which lexicographically
sorts the suffixes of the string starting at that position.
It is an influential data structure with many applications in efficient string algorithms solving problems such as exact pattern matching, repeat finding, maximum exact match (MEM) finding, document retrieval and many more.

There exist several algorithms for suffix array construction in linear time \cite{skew2003karkkainen,sais2009nong,saoverview2007puglisi} with several practical implementations \cite{louza2017inducing,libdivsufsort}.
Despite their linear time complexity, these algorithms become bottlenecks in some applications because of their linear space complexity.
This observation is especially relevant in pangenomics, where the datasets often do not fit in the computer memory.

Here, we show another crucial advantage of prefix-free graphs.
Although they do not offer any improvement in theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using them directly in subsequential computation or storing them in compressed form.
This iteration can be done without ever expanding the pangenome to its textual representation in space proportional to the sum of segment lengths and the sum of path lengths.

\subsection{Iterator preparation}
To prepare the iterator of a suffix array of the pangenome from a prefix-free graph, we need to create several data structures.
First, we concatenate all the segments into a single string using a separator $\#$ and append a sentinel $\$$.
We will call this concatenation \emph{segment join}.
An example of a segment join is in Figure \ref{fig:ids_and_positions}.

Next, we calculate the segment join's suffix array and the longest common prefix (LCP) array.
For both of these arrays, there exist algorithms with linear time and space complexity which we can use.
We note that these linear complexities are proportional to the length of the segment join, which is usually much smaller than the original pangenome.

Next, for each suffix of the segment join, we need to calculate the corresponding segment ID and position values.
The value segment ID represents in what segment the current suffix starts, and the value segment position represents at what position in that particular segment the current suffix starts.
These arrays can be computed using an inverse permutation of a suffix array of a segment join in linear time.
To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
segment ID array and segment position array.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
    \caption{
        Segment IDs and segment positions of a segment join. 
        Permuting these according to the inverse permutation of a suffix array results in segment ID array and segment position array shown in Table \ref{tab:suffix}.
    }
    \label{fig:ids_and_positions}
\end{figure}

@@

In Table \ref{tab:suffix}, one row can represent multiple positions of the
pangenome.
To identify these positions, we store some additional information in Table \ref{tab:segment}.
For each segment, we store its length, starting positions in the pangenome and the rank of the right context of these positions.
To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
Similarly to segment join, a path join is a concatenation with delimiters $\#$ and sentinel $\$$, but now constructed by concatenating the paths.
An example of a path join for our running example is in Figure \ref{fig:path_join}.
Starting positions can be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.

The computation of ranks is more involved.
It uses the normalized form of prefix-free graphs since it relies on a lexicographically smaller ID in a path representing a lexicographically smaller segment.
We construct the suffix array of the path join and find its inverse permutation $\texttt{ISA}$.
$\texttt{ISA}$ gives us ranks for each position in the path join.
To determine the rank of the right context for position $i$, we take
the value of $\texttt{ISA}[i+1]$.
Finally, we store the starts and ranks sorted by the rank values in the segment table as shown in Table \ref{tab:segment}.

@@

% the prefix-free property hold also for all suffixes of segments longer than k
% The size of prefix-free graphs can be 
% Because each segment is represented in a set $S$ only once, the sum of lengths
% of segments is usually much smaller than the length of original sequences.
